Extracting Function Signatures from Program Traces

### 2020-05-14 paper/main.tex
minor

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R has
been a significant undertaking. We had to contend with the complex design
choices of R to make it work without surprises for the end-users. We describe
two key challenges that we faced in the development of \contractr.
\begin{itemize}
\item Firstly, we faced the problem of type-checking arguments in a
  non-strict language while retaining the non-strict semantics. When a
  function is called in R, parameters are bound to unevaluated code thunks
  called promises~\cite{oopsla19}, instead of values which can be
  immediately type-checked. Naively forcing promises at function call to
  obtain a value for type-checking would violate the language semantics,
  leading to incorrect results. To preserve the original semantics,
  \contractr modifies the promise by wrapping the unevaluated promise
  expression in a call to its type-checker along with the context necessary
  to associate the promise to the function and its parameter (for failure
  messages). The contract checking happens when the promise is evaluated,
  either inside the function or many calls deep.  This works, but with one
  wrinkle. GNU R has a bytecode compiler that can sometimes optimize away
  promises; in those cases, \contractr receives values that can be
  immediately type-checked. Non-strict semantics dictate that any
  computation related to the argument should happen on the first use of the
  argument. So type-checking at this point and issuing an error message
  would violate the non-strict semantics; if this argument is never used, it
  would be incorrect to type-check it at the beginning of call. To make this
  work, \contractr mutates the function parameter binding to a promise that
  wraps the argument value in a call to the type-checker, as in the previous
  case.
\item Secondly, we faced the problem of type-checking function return
  value. In R, the result of a function call is the result of evaluating the
  last expression in the function body; so an explicit return call is absent
  from most function definitions. Thus, there could be many potential
  sub-expressions in the function body which need to be type-checked against
  the return type. To address this, \contractr registers the type-checker to
  be called on the return value through a function exit hook. This hook is
  executed in the function call's environment after it has executed. While
  this works, we have to contend with yet another wrinkle, R interpreter can
  perform longjumps in its C implementation, which causes active R function
  calls on the stack to be discarded. When they are discarded, their exit
  hooks are called, which in this case calls the registered
  type-checker. But, these functions are in the middle of an active
  computation, so they don't have a return value to type-check. \contractr
  deals with this problem by allocating a unique sentinel object which
  serves as the return value for calls that are discarded. The function call
  exit hook does not call the type-checker if this unique value happens to
  be the return value of the call.
\end{itemize}
%
%
\subsubsection{Usability}
With \contractr, we aim to provide a hassle-free path to the R developers
for adding types to untyped package code. Hence, we have paid a lot of
attention towards usability in its design. We discuss three such usability
features.
\begin{itemize}

>>>>

%
%
\subsubsection{Challenges}
Retrofitting a robust and efficient type-system as an external package in R has
been a significant undertaking. We had to contend with the complex design
choices of R to make it work without surprises for the end-users. We describe
two key challenges that we faced in the development of \contractr.
\begin{itemize}
\item Firstly, we faced the problem of type-checking arguments in a
  non-strict language while retaining the non-strict semantics. Recall that when a
  function is called in R, parameters are bound to unevaluated code thunks
  called promises~\cite{oopsla19}, instead of values which can be
  immediately type-checked. Naively forcing promises at function call to
  obtain a value for type-checking would violate the language semantics,
  leading to incorrect results. To preserve the original semantics,
  \contractr modifies the promise by wrapping the unevaluated promise
  expression in a call to its type-checker along with the context necessary
  to associate the promise to the function and its parameter (for failure
  messages). The contract checking happens when the promise is evaluated,
  either inside the function or many calls deep.  This works, but with one
  wrinkle. GNU R has a bytecode compiler that can sometimes optimize away
  promises; in those cases, \contractr receives values that can be
  immediately type-checked. Non-strict semantics dictate that any
  computation related to the argument should happen on the first use of the
  argument. So type-checking at this point and issuing an error message
  would violate the non-strict semantics; if this argument is never used, it
  would be incorrect to type-check it at the beginning of call. To make this
  work, \contractr mutates the function parameter binding to a promise that
  wraps the argument value in a call to the type-checker, as in the previous
  case.
\item Secondly, we faced the problem of type-checking function return
  value. In R, the result of a function call is the result of evaluating the
  last expression in the function body; so an explicit return call is absent
  from most function definitions. Thus, there could be many potential
  sub-expressions in the function body which need to be type-checked against
  the return type. To address this, \contractr registers the type-checker to
  be called on the return value through a function exit hook. This hook is
  executed in the function call's environment after it has executed. While
  this works, we have to contend with yet another wrinkle, R interpreter can
  perform longjumps in its C implementation, which causes active R function
  calls on the stack to be discarded. When they are discarded, their exit
  hooks are called, which in this case calls the registered
  type-checker. But, these functions are in the middle of an active
  computation, so they don't have a return value to type-check. \contractr
  deals with this problem by allocating a unique sentinel object which
  serves as the return value for calls that are discarded. The function call
  exit hook does not call the type-checker if this unique value happens to
  be the return value of the call.
\end{itemize}
%
%
\subsubsection{Usability}
With \contractr, we aim to provide a hassle-free path to the R developers
for adding types to untyped package code. Hence, we have paid a lot of
attention towards usability in its design. We discuss three such usability
features.
\begin{itemize}
